Reasonably Programmable Syntax

### 2017-05-02 popl18/syntax-popl18.tex
moved failed ICFP17 submission into a directory and copied it into popl18 for next try. made some tweaks to the abstract

\input{../../thesis/macros}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programmers, 
like pen-and-paper mathematicians, often wish to define  ``syntactic sugar'' to  lower the syntactic cost of common idioms. 
Unfortunately, the most direct mechanisms of syntactic control available to programmers, e.g. \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, do not support modular reasoning about determinism (i.e. separately defined forms can conflict syntactically with one another.) Moreover, they obscure the type and binding structure of the program. Typed hygienic term-rewriting macros (e.g. as available in Scala) are more reasonable, but they give programmers only limited syntactic control.

This paper formally introduces \emph{simple} and \emph{parametric typed syntax macros} (TSMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} at a specified type or parameterized family of types. Expansion is {strictly hygienic}, meaning that 1) expansions can refer to external bindings only via spliced terms or supplied module parameters; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
This design ensures that clients are able to reason about the binding structure of the program while holding the expansion of the program abstract -- the system needs only convey to clients the \emph{segmentation} of each literal, which gives the locations of the spliced terms. 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
We argue that this mechanism occupies a ``sweet spot'' in the design space, in that it captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. We establish these reasoning principles formally.

@@

Third-party library providers are, quite justifiably, envious of the privileged status of these otherwise seemingly ordinary standard library definitions. After all, it is not difficult to find other examples of situationally useful library-specific syntactic sugar \cite{TSLs}. For example, 1) clients of a ``collections'' library might like syntactic sugar for finite sets and dictionaries; 2) clients of a ``web programming'' library might like syntactic sugar for HTML and JSON values; 3) a compiler writer might like syntactic sugar for the terms of the object language or various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like syntactic sugar for chemical structures based on the SMILES standard \cite{anderson1987smiles}. This has motivated a long line of research into approaches that delegate more syntactic control to library providers.

@@

%% Bibliography
\bibliography{../../papers/research}

>>>>

\input{../../../thesis/macros}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Programmers, 
like pen-and-paper mathematicians, often want to define new ``syntactic sugar'' to  lower the syntactic cost of common idioms. 
Unfortunately, the most direct mechanisms of syntactic control available to programmers, e.g. \li{camlp4} \cite{ocaml-manual} and Sugar* \cite{erdweg2011sugarj,erdweg2013framework}, do not support modular reasoning about syntactic determinism -- separately defined syntax extensions can conflict syntactically with one another. Moreover, clients cannot reason abstractly, i.e. given only the program text, about types and binding. Instead, they must reason transparently, i.e. about the underlying expansion. To address these problems, this paper  formally introduces \emph{simple} and \emph{parametric typed literal macros} (TLMs), which give library providers programmatic control over both the parsing and expansion of expressions and patterns of \emph{generalized literal form} \cite{TSLs} at a specified type or parameterized family of types. Expansion is {strictly hygienic}, meaning that 1) expansions can refer to external bindings only via spliced terms or supplied module parameters; and 2)  expansions do not reveal internal bindings to spliced terms or to the remainder of the program. 
% Partial parameter application lowers the syntactic cost of this strict style. 
TLMs offer substantially more syntactic control than existing hygienic term-rewriting macro systems while ensuring that clients are able to reason modularly about syntactic determinism and abstractly about the type and binding structure of the program. The system needs only to convey to clients (e.g. via secondary notation) the \emph{segmentation} of each literal, which gives the locations of the spliced terms. 
%The system can determine this {segmentation} automatically and convey it to the client via syntax highlighting, or by presenting the client with a context-free grammar (without revealing the semantic actions associated with each production.) 
% This mechanism captures many common syntactic idioms while avoiding the problem of syntactic conflicts by construction and supplying clients with clear abstract reasoning principles. 
We specify TLMs and establish these abstract reasoning principles formally using an ML-style calculus of expressions, patterns and modules.

@@

Third-party library providers are, quite justifiably, envious of the privileged status of these otherwise ordinary standard library definitions. After all, it is not difficult to find other examples of situationally useful library-specific syntactic sugar \cite{TSLs}. For example, 1) clients of a ``collections'' library might like syntactic sugar for finite sets and dictionaries; 2) clients of a ``web programming'' library might like syntactic sugar for HTML and JSON values; 3) a compiler writer might like syntactic sugar for the terms of the object language or various intermediate languages of interest; and 4) clients of a ``chemistry'' library might like syntactic sugar for chemical structures based on the SMILES standard \cite{anderson1987smiles}. This has motivated a long line of research into approaches that delegate more syntactic control to library providers.

@@

%% Bibliography
\bibliography{../../../papers/research}
